### 2022-11-08 doc/src/IOMMU.tex
Rename MSI PTE translate-R/W mode to basic translate mode
Signed-off-by: John Hauser <31252952+jhauser-us@users.noreply.github.com>

If V~=~1 and the custom-use bit C~=~0, then bits~2:1 of the first
doubleword contain field~M (Mode).
If M~=~3, the MSI PTE specifies \emph{pass-through mode} for
accesses to the page, and if M~=~1, it specifies \emph{MRIF mode}.
Values of 0 and 2 for M are reserved.
The interpretation of an MSI PTE for each of the two defined modes is
detailed further in the next two subsections.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{MSI PTE, pass-through mode}

When an MSI PTE has fields V~=~1, C~=~0, and M~=~3
(pass-through mode), the PTE's complete format is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:10 & PPN \\
                   & bits 2:1   & M, = 3 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & ignored \\
\end{displayLinesTable}
All other bits of the first doubleword are reserved and must be set to
zeros by software.
The second doubleword is ignored by an \mbox{IOMMU} so is free for software
to use.

@@

An MSI PTE in pass-through mode allows a hypervisor to route an
MSI write intended for a virtual interrupt file to go instead to a guest
interrupt file of a real IMSIC in the machine.

@@

For RV64, the first doubleword of an MSI PTE in pass-through mode
has the same encoding as a regular {\RISCV} leaf PTE for Sv39, Sv48,
Sv39x4, or Sv48x4 page-based address translation, with PTE fields D, A,
G, U, and X all zeros and W~= R~=~1.
Hence, the MSI PTE's first doubleword appears the same as a regular
PTE that grants read and write permission (R~= W~=~1) but not execute
permissions (X~=~0).
This same-encoded regular PTE would translate an MSI write the same as
the actual MSI PTE, except that what would be the PTE's accessed (A),
dirty (D), and user (U) bits are all zeros.
An \mbox{IOMMU} needs to treat only these three bits differently for an MSI
PTE versus a regular RV64 leaf PTE.

@@

\begin{commentary}
While \mbox{IOMMU}s are expected typically to cache MSI PTEs that are
configured in pass-through mode (M~=~3), they might not cache PTEs
configured in MRIF mode (M~=~1).
Two reasons together justify not caching MSI PTEs in MRIF mode:
First, the information and actions required to store an MSI to an MRIF
are far different than normal address translation; and
second, by their nature, MSIs to MRIFs should occur less frequently.
Hence, an \mbox{IOMMU} might perform MRIF-mode processing solely as
an extension of cache-miss page table walks, leaving its address
translation cache oblivious to MRIF-mode MSI PTEs.
\end{commentary}

>>>>

If V~=~1 and the custom-use bit C~=~0, then bits~2:1 of the first
doubleword contain field~M (Mode).
If M~=~3, the MSI PTE specifies \emph{basic translate mode} for
accesses to the page, and if M~=~1, it specifies \emph{MRIF mode}.
Values of 0 and 2 for M are reserved.
The interpretation of an MSI PTE for each of the two defined modes is
detailed further in the next two subsections.

%- - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - - -
\subsection{MSI PTE, basic translate mode}

When an MSI PTE has fields V~=~1, C~=~0, and M~=~3
(basic translate mode), the PTE's complete format is:\nopagebreak
\begin{displayLinesTable}[l@{\quad}l@{\quad}l]
First doubleword:  & bit 63     & C, = 0 \\
                   & bits 53:10 & PPN \\
                   & bits 2:1   & M, = 3 \\
                   & bit 0      & V, = 1 \\
\noalign{\medskip}
Second doubleword: & ignored \\
\end{displayLinesTable}
All other bits of the first doubleword are reserved and must be set to
zeros by software.
The second doubleword is ignored by an \mbox{IOMMU} so is free for software
to use.

@@

An MSI PTE in basic translate mode allows a hypervisor to route an
MSI write intended for a virtual interrupt file to go instead to a guest
interrupt file of a real IMSIC in the machine.

@@

For RV64, the first doubleword of an MSI PTE in basic translate mode
has the same encoding as a regular {\RISCV} leaf PTE for Sv39, Sv48,
Sv39x4, or Sv48x4 page-based address translation, with PTE fields D, A,
G, U, and X all zeros and W~= R~=~1.
Hence, the MSI PTE's first doubleword appears the same as a regular
PTE that grants read and write permission (R~= W~=~1) but not execute
permissions (X~=~0).
This same-encoded regular PTE would translate an MSI write the same as
the actual MSI PTE, except that what would be the PTE's accessed (A),
dirty (D), and user (U) bits are all zeros.
An \mbox{IOMMU} needs to treat only these three bits differently for an MSI
PTE versus a regular RV64 leaf PTE.

@@

\begin{commentary}
While \mbox{IOMMU}s are expected typically to cache MSI PTEs that are
configured in basic translate mode (M~=~3), they might not cache PTEs
configured in MRIF mode (M~=~1).
Two reasons together justify not caching MSI PTEs in MRIF mode:
First, the information and actions required to store an MSI to an MRIF
are far different than normal address translation; and
second, by their nature, MSIs to MRIFs should occur less frequently.
Hence, an \mbox{IOMMU} might perform MRIF-mode processing solely as
an extension of cache-miss page table walks, leaving its address
translation cache oblivious to MRIF-mode MSI PTEs.
\end{commentary}
